---
title: "EDS 221 - Scientific programming essentials"
format: html
editor: visual
execute:
  warning: false
  message: false
toc: true
toc-depth: 4
---

# Introduction

## Terms

Programming: *writing instructions to tell computer what to do*

Data science: *"process of turning data into understanding"*

Programming language: *language that we use to communicate with computers*

Reproducibility: *Being able to reproduce outcomes of prior study* 

-   **Computational:** when information is provides about code, software, hardware and implementation details 

-   **Empirical:** when detailed info is provided about non-computational (eg. how data was collected)

-   **Statistical:** information about methods of analysis

Functions: *self contained bits of code that accept user inputs, do something with it, and return a useful output*

## Terminal Commands

**git commit -m "message"**

-   commit changes

**git push**

-   push changes to remote repository

**git pull**

-   pull to check for updates in remote repo

**mkdir** my-new-folder 

-   To create new folder 

**rmdir** my-new-file

-   Remove file from folder 

**touch**

-   To add file (eg. rmd or qmd)

**git branch**

-   Will tell you what branch you're working in

**git checkout main**

-   Will take you to branch selected, in this case **main** branch

# **Data representation and types**

## Types

#### Numeric (eg. 14.6, -0.031)

-   Integer: whole number

-   Floats, doubles (subset of numeric- more precision/more bits allotted eg. 64 bit vs 32 bit)

-   Trying to coerce numeric to integer will not work in R, in Python it will run down 

-   Combining -\> will result in the most flexible class

#### Integer (eg. 8L)

-   Whole numbers

-   Character (eg. "cat breath" or "bren")

    -   Sequence of characters create **strings**

    -   In R strings are stored as class "character"

    -   In Python stored as class "str"

#### Logical (True/False)

-   Binary and mutually exclusive values 

-   Coded as 0 and 1

#### Factors 

-   Ordered discrete levels

-   Order matters and can be specified

-   Important for ordinal variables that are included in models/analyses

## Data Structures

### Homogenous (same type)

##### 1D - Atomic vector: Flat sequence of values, all of the same type

1.  Properties of vector

    1.  Type (eg. "character" "logical")

    2.  Length

    3.  Attributes (additional associates metadata)

##### 2D- Matrix (eg. leslie matrix)

1.  Arrangement of values

2.  Usually mxn table of values

3.  Usually use dataframes

##### nD- Array

### Heterogeneous (different types)

##### 1D-List

1.  Like a vector, but can contain different data types

2.  Use list()

3.  Outputs of models are often stored as lists 

##### 2D-Data frame

1.  Can contain different classes

2.  List of equal length vectors 

3.  Make data frames manually/ import them (eg. csv, xlsx, .txt)

### Indexing: accessing elements from different data structures 

use \[ \]

-   \[row, column\]

    -   In R: first element is position = 1

    -   In Python: first element is position = 0

    -   If only give one elements (eg. \[8\]) then it refers to just column 8

# **Conditionals and logical operations, intro to for-loops**

## Boolean operators

== does this match

## Conditional Statements

#### if:

*if a condition is met, do this. Otherwise, no nothing.*

**in R**

```{r}
burrito <- 2.4 # assigning object value 

# if statement 
if (burrito > 2) {
  print("I love these!")
}
```

**in Python**

```{r}
# burrito = 2.4
# 
# if burrito > 2:
#   print("I love these!")
```

#### if-else:

*if a condition is met, do this. Otherwise, do a different thing.*

-   to extend...

    if( ) {

    } else if( ) {

    } else( )

```{r}
pika <- 89.1

if(pika > 60) {
  print("mega pika")
} else
  print("normal pika")
```

#### switch:

*allows a variable to be tested in favor of equality against a list of case values.*

```{r}
species = "mouse"

switch(species, 
       "cat" = print("Meow"),
       "dog" = print("Woof!"),
       "mouse" = print("Squeak!"))
```

### For Loops

For each country in data set...

For each csv file in directory...

for each species in data...

#### Basic For-loop

##### Strings:

```{r, results = 'hide'}
dog_names <- c("A", "B", "C", "D")

for (i in dog_names) { # this is the sequence
  print(paste("My dog's name is", i)) # this is the body of the loop; telling it what to do
}
```

##### Numeric:

```{r, results = 'hide'}
mass <- seq(from = 0, to = 12, by = 0.5)

for (i in mass) {
  new_val = i + 2
  print(new_val)
}
```

A better way to do this...because it allows you to use access information from more than 1 vector. Position based way to access elements of vectors.

```{r, results = 'hide'}
mass <- seq(from = 0, to = 12, by = 0.5)

# accessing value based on its position
  # seq_along() makes a vector of counts that is the length of chosen vector
for (i in seq_along(mass)) {
  new_val = mass[i] + 2
  print(new_val)
}

# another example
vec <- c(3, 5, 7, 8, 2)

for (i in seq_along(vec)) {
  print(vec[i] + 2)
}
```

##### Storing outputs:

-   Pre-create an empty vector to store the outcome of the for loop

```{r}
# my_output <- vector(mode = "numeric", 
#                     length = length(vector))

tigers <- c(29, 34, 82)
lions <- c(2, 18, 6)

# storage; creating an empty vector
big_cats <- vector(mode = "numeric", 
                   length = length(tigers))

for (i in seq_along(tigers)) {
  total_cats <- tigers[i] + lions[i] # indexing accessing the value in i element
  big_cats[i] <- total_cats
}
```

**Another example of for-loop:**

-   paste0() *output is a string without spaces*

```{r}

dosage <- c(10, 100, 1000)

# create empty vector
dose_index <- vector(mode = "character", 
                     length = length(dosage)) 

# fill the empty vector
for (i in seq_along(dosage)) {
  k <- paste0("zinc_", dosage[i])
  dose_index[i] <- k
}
```

#### For loops and Conditionals

-   good for groups of data/across different files (eg. if file type is csv put it *here*, if file type is html put it **there**)

```{r, results = 'hide'}
animal <- c("fi", "fy", "fo", "fum", "no")

for (i in seq_along(animal)) {
  if (animal[i] == "fi") {
    print("I love ogres!")
  } else
    print("This is the wrong castle")
}
```

#### Referring to columns in a data frame

Reference column using \$ (eg. penguins\$species)

Reference using \[ \] and calling column number

-   [ ] will call all items in column as a single data frame

-   \[\[ \]\] will call all items in column as a vector

```{r}
mean_mtcars <- vector(mode = "numeric", 
                      length = ncol(mtcars)) # make the length the # of columns in mtcars 

for (i in seq_along(mtcars)) {
  mean_val <- mean(mtcars[[i]], # return vector of values
                   na.rm = TRUE) # remove NA values so that we can still take mean
  mean_mtcars[i] <- mean_val
}
```

Iterate over a selected number/type of columns

```{r}
# data we are using 
library(palmerpenguins)
```

```{r}

for (i in seq_along(penguins)) {
  if (is.numeric(penguins[[i]])) {
    penguin_med <- median(penguins[[i]],
                          na.rm = TRUE)
    print(penguin_med)
  }
}
```

#### Nested For loops

Example with explanation in comments:

```{r, results = 'hide'}
file_prefix <- c("temp", "ph", "salinity")
file_suffix <- c(1, 2, 3, 4, 5)

for (i in seq_along(file_prefix)) { 
  for (j in seq_along(file_suffix)) {
    print(paste0(file_prefix[i], "_", file_suffix[j]))
  }
}

# starting at i = 1, 
#   has to go through every value of j before i goes to 2
#    for i = 1, output is 
#                         i1_j1, i1_j2, i1_j3, i1_j4, i1_j5
# then i = 2, output is
#   has to go through every value of j before i goes to 3
#    for i = 2, output is
#                         i2_j1, i2_j2, i2_j3, i2_j4, i2_j5
```

## Functional Programming

### apply():

*a family of functions in base R; iterating across rows or columns of a matrix*

apply(data.frame, #, function)

-   1 means iterating for columns *eg.* *apply(mtcars, 1, mean)*

-   2 means iterating for rows

lapply()

sapply()

### tapply()

*apply a function by group (factor level)*

### dplyr()::across():

*function in dplyr to apply a function across specified columns*

### {purrr}:

*R package that allows you to replace many for-loops with better code*

# Introduction to Functions

## Anatomy of a Function

Example:

```{r}
# function_name <- function(arguments) { body/outputs }

pet_count <- function(bird, dog) {
  pets = bird + dog
  return(pets)
}

# using the function
pet_count(3, 5)
```

**Another example:**

```{r}
exclaim_age <- function(age) {
  print(paste("I am", age, "years old!"))
}

# using function
exclaim_age(10)
```

## Function with Condition

```{r}
animal_age <- function(animal, age) {
  if (animal == "dog") { # if the animal inputed is dog
   print(age * 7) # then use this equation to create output
} else if (animal == "goat") { # if the animal input is goat
  print(age * 4.7) # then use this equation to create output
}
}

```

### Examples

**Example 1**

```{r}
calculate_animal_age <- function(species, age_human_yrs) {
  if (species == "dog") {
    age_human_yrs * 7
  } else if (species == "goat") {
    age_human_yrs * 4.7
  }
}

# to break it
calculate_animal_age("whale", 7)

# solution - add another else condition 
calculate_animal_age <- function(species, age_human_yrs) {
  if (species == "dog") {
    age_human_yrs * 7
  } else if (species == "goat") {
    age_human_yrs * 4.7
  } else print("Please enter goat or dog!")
}

calculate_animal_age("whale", 7)
```

**Another Example:**

Write a function that will return, given temp in Kelvin, if a user selects "metric" to the scale then return degrees C, or if they select "english" then return degrees in F.

```{r}

convert_kelvin_to_f_c <- function(temp_kelvin, scale) {
  if (scale == "metric") {
    print(temp_kelvin - 273.15)
} else if (scale == "english") {
    print((temp_kelvin - 273.15) * 9/5 + 32)
  }
}

convert_kelvin_to_f_c(50, "metric")
```

**Anotha one!!!**

Create a function that returns an awesome restaurant name given a cool animal and your favorite food.

```{r}

restaurant_name <- function(animal, food) {
  print(paste0(animal, food))
}

restaurant_name("tiger", "Apples")
```

**Lets get ittttt, one more example:**

Given temp in degrees F, write a function that converts it to Celsius.

```{r}
f_to_c <- function(f_temp) {
  print((f_temp - 39) * (5/9))
}

f_to_c(50)
```

**Just kidding, heres the last one:**

Create a function that given inputs of age and hometowns, return "I am \_\_ years old, and I'm from \_\_\_"

```{r}

home_town <- function(age, town) {
  print(paste("I am", age, "years old, and I'm from", town))
}

home_town(13, "NYC")
```

## Logistic Growth Equation

### Step 1: Plan

*I want to automate estimating the population size over a range of times, and at multiple growth rates (for a constant intitial population and carrying capacity).*

### Step 2: Pseudocode

```{r}
# out_matrix = matrix() # to sore output
# 
# for (i in sequence of growth rates) {
#   for (j in sequence of times) {
#     calc_logisitc( for i over all j)
#     store calc_population in out_matrix
#   }
# }
```

### Step 3: Create function

```{r}
logistic_growth <- function(N0, K, r, time) {
  Nt <- K / (1 + ((K-N0) / N0) * exp(-r * time))
  print(Nt)
}
```

```{r, results = 'hide'}
# create vector of times
time_vec <- seq(from = 0, to = 35, by = 0.1)

# create a sequence of growth rate values:
r_seq <- seq(from = 0.2, to = 0.4, by =0.01)

# create a Matrix to store outputs in
out_matrix <- matrix(nrow = length(time_vec),
                     ncol = length(r_seq))

for (i in 1:length(r_seq)) { # outer loop is the growth rates
  for (j in 1:length(time_vec)) { # innter loop is the time sequence values
    pop <- logistic_growth(N0 = 100, 
                           K = 6000, 
                           r = r_seq[i],
                           time = time_vec[j])
    out_matrix[j, i] <- pop # store the valye in the appropriate row and column 
  }
}
```

# Testing Functions

Unit test: *formal, automated check to make sure a piece ("unit") of your code is behaving expectantly*

-   assume you will not remember what you have tested from a function

## {testthat} Functions

*Makes it easy to describe what you expect a function to do*

-   expect_equal() *does the function return the expected values?*

-   expect_type() *does the function return the expected type?*

-   expect_length() *if a vector is returned, is it the expected length?*

-   expect_true() *does the function return TRUE?*

Example:

-   If true, nothing is returned. If the function does not pass a test, an error message will be returned.

```{r}
# this '1' value at the end is the expected length 
## expect_length(fish_mass(alpha = 1, beta = 2, fish_length = 3), 1)

```

# Finding, exploring, and using packages

## Finding useful packages

-   Check out [this](https://cran.r-project.org/web/views/) link to see what packages are available to use on CRAN. And [this](https://ropensci.org/) link to check out R Open Science packages.

## Vetting Packages Checklist

-   Maturity (how long its been around for)

    -   CRAN old sources -\> see how many iterations it has gone through

    -   look at lifecycle

-   Active development (are still people contributing? modifying for updates?)

    -   look at commits, issues, added features

-   Recently updated

    -   commits, closed issues

-   Documentation available

-   Used and published in similar projects

-   License

-   Semantics intuitive

-   Functions that get the job done

-   Arguments to support your needs

-   Dependencies reported/reasonable (packages your code is using - try not to use too many)

## Citing Packages

Citation function to cite the packages used!

-   citation("package_name")
